From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-03-06 paper/paper.Rmd
tweaks

## Use trends style to quickly count refs
csl: trendsjournals.csl
#csl: ecology-letters.csl
bibliography: refs.bib
output: 
  rticles::elsevier_article:
    keep_tex: true
layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \usepackage[nomarkers]{endfloat}
  \journal{Ecology Letters}
  \linenumbers
  \usepackage{setspace}
  \doublespacing

abstract: |
  # Article Type: Review & Synthesis
  # Running title: Noise phenomena
  # Figures: 4
  # Tables: 0
  # Word count: 
  # Word count in abstract: 
  # Number of References: 117
  # Archive statement:

@@

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5, cache = FALSE)
library(tidyverse)
library(gridExtra)
library(ggthemes)
library(png)
library(scales)
colours <- ptol_pal()(2)

@@

To the empirical ecologist, stochasticity, or more simply, "noise,"
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of statistical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
from thinking of stochasticity in such terms; where noise is a nuisance
that obscures the "deterministic skeleton" of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena [@Coulson2004].  Yet this transition from "noise the nuisance"
to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
relatively little impact in empirical thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become "noise the informer."  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empirical communities by 
seeing noise not as mathematical curiosity or statistical bugbear, but as a
source for new opportunities for inference.

@@

To emphasize the underlying trend in the changing roles in which we
see and understand noisy processes, I will also restrict my focus 
to relatively simple models primarily from population ecology context.
Simplicity not only makes examples (in equations and in code) more
tractable but also allows us to focus on aspects that are germane to many
contexts rather than unique to particular complexities [@Levins1966; @Bartlett1960].
Nevertheless, that complexity matters -- few themes have been better 
emphasized in the theoretical literature [@Bjornstad2001].  Both
the foundational literature and recent research continue to echo the
theme of understanding the impact different real world complexities
have in stochastic dynamics, including the importance
of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989; @Tuljapurkar1997;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g.
@Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010], Individual
heterogeneity [e.g. @Vindenes2015; @Hart2016] or the coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010;
@Lee2017; @Spanio2017] or periodic [@Coulson2001; @Bjornstad2001;
@Keeling2001] structure in environmental noise, or the interaction
of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015].  As such, we will 
rely on both textbooks and recent reviews to provide a proper treatment
of these issues, and focus on broader trends that may be less apparent
to non-specialists.

This review is structured into three sections: Origins of noise, emergent
phenomena, and noise-driven inference.  The first section lays the conceptual
groundwork we will need, while also highlighting a shift to more and more
mechanistically rooted descriptions of noise.  We will see where the common
formulation of 'deterministic skeleton plus noise term' comes from, how it 
is best justified, and when it breaks down.  The second section introduces
"noise the creator," showing examples of ecological phenomena generated by
stochasticity.  These examples will be familiar to most specialists but
illustrate a different way of thinking than held by most ecologists, where
noise is only a nuisance to be filtered or averaged out.  The third and
final section, "noise the informer," turns these examples back-to-front, 
asking what noise can tell us about a system: such as its underlying resilience
or stability, or the approach of a catastrophic shift. Examples are fewer
here, and have largely yet to benefit from the introduction of either
the rigorous theorems or more complex models so plentiful in the previous
sections.  Yet the promise of prediction, of early warning signs before
tipping points, have spurred broad interest in such noise-based inference.
This review is a call to both deepen the connection to mechanism and better
formalize this thinking, but also look more broadly into other ways in which
noisy phenomena can help inform and predict underlying processes.

<!--
I approach this synthesis in three parts.  First, I will sketch the ground-work
for the origins of noise in any model as statistical summary of large number of
additional processes from some lower level that are modeled only in aggregate.
This will provide the fundamentals we need to illustrate several key examples
of noise driven phenomena in ecological dynamics: noise the creator.  Finally,
we will be able to invert some of these examples; instead of going from process
to pattern, we will go from pattern to process: noise the informer.  
-->

@@

What is noise? Where does it come from? 
Many stochastic models are not explicit about whether the noise terms
they introduce are intended to reflect intrinsic or extrinsic factors.
Early quantitative ecological models first formulated at the population
level in deterministic settings. The notion of a deterministic
"skeleton" -- responsible for the phenomena of interest -- that is obscured
from view by the nuisance of noise has slowly given way to recognition
that stochastic and deterministic elements not only interact in 
interesting ways [@Coulson2004] but also share a common origin [e.g. @Black2012].
A central theme of stochastic modeling research over the past
several decades has been an emphasis on making this relationship
between stochasticity and underlying processes more precise:
that is, on the origins of noise.

Simply put, noise is all that we leave out of our model. 
Whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
This literature has largely divided the origins of stochasticity between 
*intrinsic* factors: where stochasticity is introduced to reflect
the reality that our variables of interest are intrinsically 
statistical averages of lower-level processes, and *extrinsic*
factors: where stochasticity reflects the impact of other variables
we are not explicitly modeling at all [e.g. @Keeling2012].  **Demographic stochasticity**
is the primary (though not the only) example of an intrinsic factor,
which reflects the fact that the continuous variable we often think of a
"population size" or "population density" is an average over a 
lower level process of births and deaths among a finite number of 
individuals [@Melbourne2012].  **Environmental stochasticity** is the primary example 
of an extrinsic factor, reflecting the fact that parameters we may
treat as constants in a model in fact depend on variables such as 
temperature or soil moisture that are not explicitly modeled by our 
equations, but simplified into statistical summaries of those processes
[@Ripa2012].

It is often suggested that stochasticity is "inherent" or "fundamental,"
e.g. "demographic stochasticity arises because the birth of an individual is probabilistic"
[@Melbourne2012]. It is easy to forget that this is merely an assertion about a 
model, not an assertion about reality [@Bartlett1960].  Only in quantum mechanics do we find 
inherently probabilistic properties: births and deaths of organisms all have
far more mechanistic explanations which we simply summarize in statistical terms:
on average -- that is, summing over a wide range of possible mechanisms we have chosen
not to explicitly model -- an individual death occurs at rate $\lambda$.  
This is in fact an extrinsic factor (variables not modeled explicitly),
which makes birth appear probabilistic; (just as we describe the toss of a coin
as probabilistic when in fact it is a mechanistic outcome of Newton's laws).  
Demographic stochasticity arises only subsequently, when we change scales from
a description at the individual level to one at the population level.

In the extensive literature developing the consequences of both demographic
and environmental noise in models of increasing complexity, it can 
be easy to miss the equally important developments in making the origins
of these sources of noise more precise. Models in which stochasticity was
essentially "tacked on" to a deterministic skeleton can be substituted for 
approaches that permit both the skeleton and noise to emerge from underlying
processes.  This more mechanistically specific approach to the origins of 
noise can indeed recover the classic formulations, but it can also 
help highlight where those models break down.
Much of the literature [e.g. see @Ovaskainen2010] has focused on the following
model, established in highly influential papers of @Leigh1981, @Lande1993,
@Foley1994:

@@

Demographic stochasticity refers to fluctuations in population sizes
or densities that arise from the fundamentally discrete nature of
individual birth and death events.  Demographic stochasticity is a
particularly instructive case for illustrating a mechanism for how
noise arises as an aggregate description from a lower-level mechanistic
process.  We summarize the myriad lower-level processes that
mechanistically lead to the event of a 'birth' in the population
as a probability: In a population of $N$ identical individuals at
time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate
that can depend on the population size, $N$), which increases the
population size to $N+1$. Similarly, death events occur with probability
$d_t(N_t)$, decreasing the population size by one individual, to $N-1$.  Assuming
each of these events are independent, this is a state-dependent
Poisson process. The change in the probability of being in state
$N$ is given by the sum over the ways to enter the state, minus the
ways to leave the state: a simple expression of probability balance
known as the "master equation" [@vanKampen2007].  Note that in general
this approach is equally applicable to stochastic transitions of any sort,
not just step sizes of +/- 1 and not just birth and death events, but
can include transitions between stage classes or trait values, including
mutations to continuously-valued traits in evolutionary dynamics [e.g. @Boettiger2010].

@@

van Kampen's approximation follows a central-limit theorem style argument established rigorously in
the theorems of Kurtz at about the same time [@Kurtz1970 & @Kurtz1971, @Kurtz1972, in particular, see Theorem 4.4. @Kurtz1978].
The heart of the approximation is to imagine a population on some large landscape of which
we sample only a fixed area to estimate the population density.  That area is our 
system size, which as it expands includes more and more total individuals and
experiences proportionally smaller fluctuations in population density.  This 
results in the conclusion that the population density obeys a simple Gaussian
distribution, whose mean size $x$ and variance  $\sigma^2$ are given by *ordinary* differential equations:

@@

This result is striking for a variety of reasons.  First, we see the recovery of the 
"deterministic skeleton," in which we naively replace population density $x/N$ for
population count $n$ describes the *expected* density of the population
(i.e. averaged over an ensemble of replicate realizations of the Gillespie simulations),
and with remarkable accuracy, up to order $1/N$, i.e. a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
The variance term is equally interesting.  First, we note that noise arises from
the interaction of individual birth and death events that defined our process,
illustrating the origins of this noise.  Because both individual events of births
and deaths add randomness, they both add to the growth of variation observed, i.e.
$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma^2$ are *non-linear* whenever
$b(x)$ or $d(x)$ are non-linear, so that different values of the population density
will experience different noise intensities! In fact, this variance can only
have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] < 0$)
Third, this expansion shows the variance
equation is also less precise, accurate only to order $1/\sqrt{N}$. 
This variance equation will be important to us later as we examine the relative
roles of noise and non-linearity in noise-driven phenomena such as stochastic
transitions between alternate stable states.

@@

$$\langle x_0, x_{t} \rangle = \frac{b(x) + d(x)}{2 \partial_x \left[ b(x) - d(x) \right]} \exp\lbrace  -t \left| \partial_x b(x) - \partial_x d(x) \right|  \rbrace $$
which will also prove useful later on.  Note this expression simply states that the equilibrium
variance is a balance between noise terms, $b(x)+ d(x)$ and the slope or eigenvalue of the
deterministic skeleton, $2 \partial_x \left[ b(x) - d(x) \right]$, (a balance that forms the basis of the
the fluctuation-dissipation theorem) and that correlations fall off exponentially at a rate given by this slope.   
We will see both auto-correlation and variance play similar roles to infer this eigenvalue when we 
reach the role of noise as informer.

In the interests of a theoretical synthesis, I observe that this result
is *different* than the canonical equation, Eq \eqref{canonical}, or more generally
the diffusion approximation [or Kramer-Moyal expansion @Kramers1940; @Moyal1949; @Gardiner2009]
which is much more typical in the ecological literature [e.g. the books of @Nisbet1982 and @Lande2003
or the review in @Ovaskainen2010]. @Black2012 gives an excellent introduction 
to the van Kampen expansion as a connection between individual-based and population-level
model descriptions and applications, but fails to point out these differences,
mistakenly characterizing the van Kampen expansion as synonymous with the diffusion approximation.
In brief, the diffusion approximation does not lead to a Gaussian distribution
(equivalently, an SDE or PDE with linear noise term, see Appendix B), but rather
a *non-linear* SDE of the form (e.g. @Ovaskainen2010)

@@

The van Kampen expansion can also be useful in illustrating how extrinsic
factors such as environmental noise enter into a stochastic model.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly auto-correlated noise 
(see derivation, Appendix B).  Auto-correlated or "colored" environmental
noise has been a topic of significant theoretical interest 
[e.g. @Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015], but such 
formulations start from the empirical observation of auto-correlation 
in environmental time series rather than a process-based derivation.
As the expansion clearly shows, any bounded continuous-time process will
involve auto-correlation -- measure a continuous process at ever-closer intervals in
time and these measurements must converge. As the derivation in Appendix B makes clear,
it is rather the the *relative* time scales of the environmental process
and population level process that really matter (more precisely, the relative 
magnitudes of the auto-correlation time of the environmental dynamics and population dynamics.)
An environment with comparatively short auto-correlation scale will act
effectively as white noise upon the population dynamics.

@@

$$\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e - c)^2} \sigma_e^2$$

@@

These examples of how a stochastic model arises from both intrinsic and extrinsic
sources of noise through the use of the system size expansion serve both
as critique and justification of the familiar notion of a stochastic model as 
a deterministic skeleton plus some Gaussian white noise term.  Critique, because
the expansions show that both "noise" and "deterministic skeleton" are the 
result of lower-level mechanistic processes, whose precise form will 
depend on the details of those processes.  Yet the expansion also shows us that
this Gaussian noise description can be a surprisingly good approximation,
that noise intensity will converge to a stationary value whenever the system
dynamics have a stable point, and that the contribution of extrinsic
(i.e. environmental) noise factors will dominate in large systems.  These
justifications will allow us to pursue our analysis of noise-driven phenomena
using simple models in which a nonlinear deterministic skeleton is driven by
Gaussian environmental noise.

<!-- This can be condensed or dropped to save space -->

## Origins of other forms of stochasticity

Stochasticity frequently enters our models in other ways as well.
For instance, most population-level models treat all individuals
within the population, or at least within a particular age or stage
class, as identical.  *Individual heterogeneity* in relevant traits
(growth rates, dispersal ability) can be represented probabilistically
as a distribution rather than a fixed value, making it function as
another form of random variation in a model [@Coulson2001;
@Schreiber2010; @Vindenes2015].  *Spatial heterogeneity* can likewise be reflected
in stochastic parameters of a spatially implicit model, or as
spatially varying environmental stochasticity in a spatially explicit
model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
interaction between individual or spatial variation and other sources
of noise can drive patterns including the potential for persistence,
coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
@Schreiber2017].

*Observational error* acts
much like another form of stochasticity, arising from sampling
effects or limitations of instruments, etc. Yet unlike the sources
described above, this uncertainty does not impact the actual state
of our system, and thus does not give rise to any additional phenomena
or information about that process, though it can obscure it
[@Knape2011; @deValpine2002].
*Deterministic chaos* [@Desharnais2001] is another mechanism of introducing
significant variation into a simple model. Chaos can a look a lot
like noise: indeed, it can be observationally indistinguishable
from a stochastic process [@Ellner2005; @Werndl2009].  Despite
substantial interest in previous decades and the relative ease of
encountering chaos in simple ecological models (particularly in
differences equations), and convincing reconstructions of chaotic
dynamics in laboratory manipulations [@Costantino1997] the prevalence
of chaotic dynamics as a driving factor in most ecological systems
remains unclear. After all, the great variation of natural populations
is hardly surprising: the undeniable influences of variation created
by all those things left out of our models, all that individual
detail from intrinsic, lower level processes and external environmental
processes that becomes aggregated in our simple models all but begs
for a stochastic description.

@@

The loss of a species is without a doubt one of the most important
possible consequences of noise. Unsurprisingly few topics have
received greater attention then the question of the persistence of
a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2010] and coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017]
in stochastic models.  Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing auto-correlated noise [e.g. @Ripa; @Ripa1999; 
@Petchey, @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues.

@@

<!--

Insufficient space left for any treatment of these...
Permenance, robust permenance

**Bet hedging**, @Lewontin1969 & Cohen 1969; Levins 1969;

-->

@@

$$x_{t+1} = x_t + x_t r \left(1 - \frac{x_t}{K}\right) - b x_t  y_t + \xi_{x,t}$$
where $x_t$ is the prey density at time $t$, $r$ the growth rate, $K$ the carrying capacity, $b$ an encounter rate with the predator $y$, who obeys the equation:

$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t}$$

>>>>

## Use trends style to quickly count refs
#csl: trendsjournals.csl
csl: ecology-letters.csl
bibliography: refs.bib
output: 
  rticles::elsevier_article:
    keep_tex: true
layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \usepackage[nomarkers]{endfloat}
  \journal{Ecology Letters}
  \linenumbers
  \usepackage{setspace}
  \doublespacing

abstract: |
  # Article Type: Review & Synthesis
  # Running title: From noise to knowledge
  # Figures: 4
  # Tables: 0
  # Word count: 7430
  # Word count in abstract: 181
  # Number of References: 117
  # Archive statement:

@@

```{r libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="cairo_pdf", fig.width=7, fig.height=3.5, cache = TRUE)
library(tidyverse)
library(gridExtra)
library(ggthemes)
library(png)
library(scales)
colours <- ptol_pal()(2)

@@

To the empirical ecologist, stochasticity, or more simply, noise,
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of statistical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
from thinking of stochasticity in such terms; where noise is a nuisance
that obscures the deterministic skeleton of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena [@Coulson2004].  Yet this transition from *noise the nuisance*
to *noise the creator* of ecological phenomena has had, with a few notable exceptions,
relatively little impact in empirical thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become *noise the informer*.  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empirical communities by 
seeing noise not as mathematical curiosity or statistical bugbear, but as a
source for new opportunities for inference.

@@

To emphasize the underlying trend in the changing roles in which we
see and understand noisy processes, I will also restrict my focus 
to relatively simple models primarily from population ecology context.
Simplicity not only makes examples (in equations and in code) more
tractable but also allows us to focus on aspects that are germane to many
contexts rather than unique to particular complexities [@Levins1966; @Bartlett1960].
Nevertheless, that complexity matters -- few themes have been better 
emphasized in the theoretical literature [@Bjornstad2001].  Both
the foundational literature and recent research continue to echo the
theme of understanding the impact different real world complexities
have in stochastic dynamics, including the importance
of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989; @Tuljapurkar1997;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g.
@Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010], Individual
heterogeneity [e.g. @Vindenes2015; @Hart2016] or the coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007;
@Schreiber2017]; correlated [@Ripa; @Ripa1999; @Petchey; @Schreiber2010;
@Lee2017; @Spanio2017] or periodic [@Coulson2001; @Bjornstad2001;
@Keeling2001] structure in environmental noise, or the interaction
of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015].  As such, we will 
rely on both textbooks and recent reviews to provide a proper treatment
of these issues, and focus on broader trends that may be less apparent
to non-specialists.

This review is structured into three sections: Origins of noise, emergent
phenomena, and noise-driven inference.  The first section lays the conceptual
groundwork we will need, while also highlighting a shift to more and more
mechanistically rooted descriptions of noise.  We will see where the common
formulation of "deterministic skeleton plus noise term" comes from, how it 
is best justified, and when it breaks down.  The second section introduces
*noise the creator*, showing examples of ecological phenomena generated by
stochasticity.  These examples will be familiar to most specialists but
illustrate a different way of thinking than held by most ecologists, where
noise is only a nuisance to be filtered or averaged out.  The third and
final section, *noise the informer*, turns these examples back-to-front, 
asking what noise can tell us about a system: such as its underlying resilience
or stability, or the approach of a catastrophic shift. Examples are fewer
here, and have largely yet to benefit from the introduction of either
the rigorous theorems or more complex models so plentiful in the previous
sections.  Yet the promise of prediction, of early warning signs before
tipping points, have spurred broad interest in such noise-based inference.
This review is a call to both deepen the connection to mechanism and better
formalize this thinking, but also look more broadly into other ways in which
noisy phenomena can help inform and predict underlying processes.

@@

What is noise? Where does it come from? 
Many stochastic models are not explicit about whether the noise terms
they introduce are intended to reflect intrinsic or extrinsic factors.
Early quantitative ecological models first formulated at the population
level in deterministic settings. The notion of a deterministic
skeleton -- responsible for the phenomena of interest -- that is obscured
from view by the nuisance of noise has slowly given way to recognition
that stochastic and deterministic elements not only interact in 
interesting ways [@Coulson2004] but also share a common origin [e.g. @Black2012].
A central theme of stochastic modeling research over the past
several decades has been an emphasis on making this relationship
between stochasticity and underlying processes more precise:
that is, on the origins of noise.

Simply put, noise is all that we leave out of our model. 
Whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
This literature has largely divided the origins of stochasticity between 
*intrinsic* factors: where stochasticity is introduced to reflect
the reality that our variables of interest are intrinsically 
statistical averages of lower-level processes, and *extrinsic*
factors: where stochasticity reflects the impact of other variables
we are not explicitly modeling at all [e.g. @Keeling2012].  **Demographic stochasticity**
is the primary (though not the only) example of an intrinsic factor,
which reflects the fact that the continuous variable we often think of a
population size or population density is an average over a 
lower level process of births and deaths among a finite number of 
individuals [@Melbourne2012].  **Environmental stochasticity** is the primary example 
of an extrinsic factor, reflecting the fact that parameters we may
treat as constants in a model in fact depend on variables such as 
temperature or soil moisture that are not explicitly modeled by our 
equations, but simplified into statistical summaries of those processes
[@Ripa2012].

It is often suggested that stochasticity is inherent or fundamental,
e.g. demographic stochasticity arises because the birth of an individual is probabilistic
[@Melbourne2012]. It is easy to forget that this is merely an assertion about a 
model, not an assertion about reality [@Bartlett1960].  Only in quantum mechanics do we find 
inherently probabilistic properties: births and deaths of organisms all have
far more mechanistic explanations which we simply summarize in statistical terms:
on average -- that is, summing over a wide range of possible mechanisms we have chosen
not to explicitly model -- an individual death occurs at rate $\lambda$.  
This is in fact an extrinsic factor (variables not modeled explicitly),
which makes birth appear probabilistic; (just as we describe the toss of a coin
as probabilistic when in fact it is a mechanistic outcome of Newton's laws).  
Demographic stochasticity arises only subsequently, when we change scales from
a description at the individual level to one at the population level.

In the extensive literature developing the consequences of both demographic
and environmental noise in models of increasing complexity, it can 
be easy to miss the equally important developments in making the origins
of these sources of noise more precise. Models in which stochasticity was
essentially tacked on to a deterministic skeleton can be substituted for 
approaches that permit both the skeleton and noise to emerge from underlying
processes.  This more mechanistically specific approach to the origins of 
noise can indeed recover the classic formulations, but it can also 
help highlight where those models break down.
Much of the literature [e.g. see @Ovaskainen2010] has focused on the following
model, established in highly influential papers of @Leigh1981, @Lande1993,
@Foley1994:

@@

Demographic stochasticity refers to fluctuations in population sizes
or densities that arise from the fundamentally discrete nature of
individual birth and death events.  Demographic stochasticity is a
particularly instructive case for illustrating a mechanism for how
noise arises as an aggregate description from a lower-level mechanistic
process.  We summarize the myriad lower-level processes that
mechanistically lead to the event of a 'birth' in the population
as a probability: In a population of $N$ identical individuals at
time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate
that can depend on the population size, $N$), which increases the
population size to $N+1$. Similarly, death events occur with probability
$d_t(N_t)$, decreasing the population size by one individual, to $N-1$.  Assuming
each of these events are independent, this is a state-dependent
Poisson process. The change in the probability of being in state
$N$ is given by the sum over the ways to enter the state, minus the
ways to leave the state: a simple expression of probability balance
known as the master equation [@vanKampen2007].  Note that in general
this approach is equally applicable to stochastic transitions of any sort,
not just step sizes of +/- 1 and not just birth and death events, but
can include transitions between stage classes or trait values, including
mutations to continuously-valued traits in evolutionary dynamics [e.g. @Boettiger2010].

@@

van Kampen's approximation follows a central-limit theorem style argument established rigorously in
the theorems of Kurtz at about the same time [e.g @Kurtz1970; @Kurtz1972; @Kurtz1978; theorem 4.4.].
The heart of the approximation is to imagine a population on some large landscape of which
we sample only a fixed area to estimate the population density.  That area is our 
system size, which as it expands includes more and more total individuals and
experiences proportionally smaller fluctuations in population density.  This 
results in the conclusion that the population density obeys a simple Gaussian
distribution, whose mean size $x$ and variance  $\sigma^2$ are given by *ordinary* differential equations:

@@

This result is striking for a variety of reasons.  First, we see the recovery of the 
deterministic skeleton, in which we naively replace population density $x/N$ for
population count $n$ describes the *expected* density of the population
(i.e. averaged over an ensemble of replicate realizations of the Gillespie simulations),
and with remarkable accuracy, up to order $1/N$, i.e. a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
The variance term is equally interesting.  First, we note that noise arises from
the interaction of individual birth and death events that defined our process,
illustrating the origins of this noise.  Because both individual events of births
and deaths add randomness, they both add to the growth of variation observed, i.e.
$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma^2$ are *non-linear* whenever
$b(x)$ or $d(x)$ are non-linear, so that different values of the population density
will experience different noise intensities! In fact, this variance can only
have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] < 0$)
Third, this expansion shows the variance
equation is also less precise, accurate only to order $1/\sqrt{N}$. 
This variance equation will be important to us later as we examine the relative
roles of noise and non-linearity in noise-driven phenomena such as stochastic
transitions between alternate stable states.

@@

\begin{equation}
\langle x_0, x_{t} \rangle = \frac{b(x) + d(x)}{2 \partial_x \left[ b(x) - d(x) \right]} \exp\lbrace  -t \left| \partial_x b(x) - \partial_x d(x) \right|  \rbrace \label{autocor}
\end{equation}

which will also prove useful later on.  Note this expression simply states that the equilibrium
variance is a balance between noise terms, $b(x)+ d(x)$ and the slope or eigenvalue of the
deterministic skeleton, $2 \partial_x \left[ b(x) - d(x) \right]$, (a balance that forms the basis of the
the fluctuation-dissipation theorem) and that correlations fall off exponentially at a rate given by this slope.   
We will see both auto-correlation and variance play similar roles to infer this eigenvalue when we 
reach the role of noise as informer.

In the interests of a theoretical synthesis, I observe that this result
is *different* than the canonical equation, Eq \eqref{canonical}, or more generally
the diffusion approximation [or Kramer-Moyal expansion @Kramers1940; @Moyal1949; @Gardiner2009]
which is much more typical in the ecological literature [e.g. @Nisbet1982; @Lande2003; 
@Ovaskainen2010]. @Black2012 gives an excellent introduction 
to the van Kampen expansion as a connection between individual-based and population-level
model descriptions and applications, but fails to point out these differences,
mistakenly characterizing the van Kampen expansion as synonymous with the diffusion approximation.
In brief, the diffusion approximation does not lead to a Gaussian distribution
(equivalently, an SDE or PDE with linear noise term, see Appendix B), but rather
a *non-linear* SDE of the form (e.g. @Ovaskainen2010)

@@

The van Kampen expansion can also be useful in illustrating how extrinsic
factors such as environmental noise enter into a stochastic model.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly auto-correlated noise 
(see derivation, Appendix B).  Auto-correlated or colored environmental
noise has been a topic of significant theoretical interest 
[e.g. @Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015], but such 
formulations start from the empirical observation of auto-correlation 
in environmental time series rather than a process-based derivation.
As the expansion clearly shows, any bounded continuous-time process will
involve auto-correlation -- measure a continuous process at ever-closer intervals in
time and these measurements must converge. As the derivation in Appendix B makes clear,
it is rather the the *relative* time scales of the environmental process
and population level process that really matter (more precisely, the relative 
magnitudes of the auto-correlation time of the environmental dynamics and population dynamics.)
An environment with comparatively short auto-correlation scale will act
effectively as white noise upon the population dynamics.

@@

\begin{equation}
\sigma_n^2 = \frac{\bar e}{c}N + \frac{\left[(1-\frac{\bar{e}}{c}) N\right]^2}{(\bar e - c)^2} \sigma_e^2 \label{env-noise}
\end{equation}

@@

These examples of how a stochastic model arises from both intrinsic and extrinsic
sources of noise through the use of the system size expansion serve both
as critique and justification of the familiar notion of a stochastic model as 
a deterministic skeleton plus some Gaussian white noise term.  Critique, because
the expansions show that both noise and deterministic skeleton are the 
result of lower-level mechanistic processes, whose precise form will 
depend on the details of those processes.  Yet the expansion also shows us that
this Gaussian noise description can be a surprisingly good approximation,
that noise intensity will converge to a stationary value whenever the system
dynamics have a stable point, and that the contribution of extrinsic
(i.e. environmental) noise factors will dominate in large systems.  These
justifications will allow us to pursue our analysis of noise-driven phenomena
using simple models in which a nonlinear deterministic skeleton is driven by
Gaussian environmental noise.

@@

The loss of a species is without a doubt one of the most important
possible consequences of noise. Unsurprisingly few topics have
received greater attention then the question of the persistence of
a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2010] and coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017]
in stochastic models.  Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing auto-correlated noise [e.g. @Ripa; @Ripa1999; 
@Petchey; @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Hastings1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues.

@@

@@

\begin{equation}
x_{t+1} = x_t + x_t r \left(1 - \frac{x_t}{K}\right) - b x_t  y_t + \xi_{x,t} \label{quasi-cycle-prey}
\end{equation}

where $x_t$ is the prey density at time $t$, $r$ the growth rate, $K$ the carrying capacity, $b$ an encounter rate with the predator $y$, who obeys the equation:

\begin{equation}
y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t} \label{quasi-cycle-predator}
\end{equation}
